An exposition on the propriety of restricted Boltzmann machines

### 2016-09-05 writing/draft.Rmd
started on instability section

The data mining and machine learning communities have recently found a great interest in deep learning, specifically in stacked restricted Boltzmann machines (RBMs). (**TODO: deep learning, deep RBM references.**) A RBM is an undirected probabilistic graphical model (for discrete or continuous random variables) with two layers, one hidden and one visible, with no conditional dependency within a layer (**TODO: RBM references.**). These models have reportedly been used with success in classification of images. (**TODO: classification references.**) However, the model properties are largely unexplored in the literature and the commonly cited fitting methodology remains heuristic-based and abstruse. In this paper, we attempt to provide steps toward a thorough understanding of the model and its behavior from the perspective of statistical model theoryand then explore the possibility of a rigorous fitting methodology. We find the RBM model class to be deficient in two fundamental ways.

@@

### Instability
@schweinberger2011instability introduces a related concept, *instability*, which occurs when small changes in the natural parameters result in large changes in probability masses. This can be characterized of as excessive sensitivity in the model, where a small change in $\boldsymbol x$ leads to a large change in the probability mass function. @schweinberger2011instability provides the following tool for detecting instability in the a discrete exponential family model.

@@

Within the context of the RBM, this definition concerns the joint distribution of both hidden and visible nodes, where $N$ is the number of nodes in the model, $H + V$. However, instability in terms of only the visible nodes is a more useful measure of model impropriety to be concerned with. To see this, consider the example of classifying images using an RBM. One would be interested in the maximum change in the likelihood when a single pixel is changed in the visibles as a proxy for instability in the model. Thus, we can instead concern ourselves with the notion of instability that deals with the maximum likelihood ratio of two images that differ in only one pixel.

@@

Let $R(\boldsymbol \theta) \equiv \max_{\boldsymbol v} \max_{\boldsymbol h}Q(\boldsymbol x) - \min_{\boldsymbol v}\max_{\boldsymbol h}Q(\boldsymbol x) - H\log 2$. Then, if $R(\boldsymbol \theta) > C*V$, this implies that the maximum log-likelihood ratio of two images that differ in only one pixel is greater than $C$. So, by examining $R(\boldsymbol\theta)/V$, we can determine a lower bound for the maximum log-likelihood ratio of two images that differ in only one pixel, which if large is consistent with the definition of instability detailed above.

Instability in a model is in itself a measure of model impropriety, however there is also a relationship between instability and near-degeneracy.

>>>>

The data mining and machine learning communities have recently found a great interest in deep learning, specifically in stacked restricted Boltzmann machines (RBMs). (**TODO: deep learning, deep RBM references.**) A RBM is an undirected probabilistic graphical model (for discrete or continuous random variables) with two layers, one hidden and one visible, with no conditional dependency within a layer (**TODO: RBM references.**). These models have reportedly been used with success in classification of images. (**TODO: classification references.**) However, the model properties are largely unexplored in the literature and the commonly cited fitting methodology remains heuristic-based and abstruse. In this paper, we attempt to provide steps toward a thorough understanding of the model and its behavior from the perspective of statistical model theory and then explore the possibility of a rigorous fitting methodology. We find the RBM model class to be deficient in two fundamental ways.

@@

### Instability

@schweinberger2011instability introduces a related concept, *instability*, which means [**TODO: that small changes in the natural parameters result in large changes in probability masses**]. This is characterized by excessive sensitivity in the model, where a small change in $\boldsymbol x$ leads to a large change in the probability mass function. @schweinberger2011instability provides the following formal definition for instability in a discrete exponential family model where [**TODO: notation goes here**].

@@

For RBMs, an alaogous concept could concern the joint distribution of both hidden and visible nodes, where $N$ is the number of nodes in the model, $H + V$. However, instability in terms of only the visible nodes is a more useful measure of model impropriety. To see this, consider modeling images using an RBM. One would be interested in the maximum change in the likelihood when a single pixel is changed in the visibles as a proxy for instability in the model. So we concern ourselves with a notion of instability that deals with the maximum likelihood ratio of two vectors of visibless that differ at only one coordinate.

@@

Let $R(\boldsymbol \theta) \equiv \max_{\boldsymbol v} \max_{\boldsymbol h}Q(\boldsymbol x) - \min_{\boldsymbol v}\max_{\boldsymbol h}Q(\boldsymbol x) - H\log 2$. Then, if $R(\boldsymbol \theta) > C*V$, this implies that the maximum log-likelihood ratio of two observations that differ in only one element is greater than $C$. So, by examining $R(\boldsymbol\theta)/V$, we can determine a lower bound for the maximum log-likelihood ratio of two sets of observables that differ in only one element, which if large is consistent with the concept of instability detailed above.

Instability in a model is in itself a measure of model impropriety, however there is also a relationship between instability and near-degeneracy in a discrete exponental model.
